Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-21 2015-turklang-kazdep/paper.tex
some bibfixes, reformatted tree stuff, etc. and more stuff since then argh
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60931 72bbbca6-d526-0410-a7d9-f06f51895060

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
\newcommand{\gmk}[1]{{\rm {\sc \texttt{#1}}}}
\newcommand{\kazakh}[1]{{\em #1}}
\newcommand{\gloss}[1]{`#1'}
\newcommand{\tag}[1]{\texttt{‹#1›}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{{\ll \textsc{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
\newcommand{\tilda}{{\fsfont ∼}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\begin{table}[htbp]
	\centering
	\caption{The default (first line of each) and ``zero-derived'' uses of nominals, adjectives, and adverbs}\label{tab:zeroderiv}
	\begin{tabular}{lp{8em}p{22em}}
		\toprule
			\textbf{Category} & \textbf{Function} & \textbf{Example} \\
		\midrule
			Nominals & Substantive\newline
							Attributive\newline
							Adverbial
						& hargle\newline 
							jargal\newline
							бір рет \gloss{once} \\
			Adjectives & Attributive\newline
							Substantive\newline
							Adverbial
						& foo \newline
						\newline
						\\
			Adverbs & Adverbial
						& 
						\\
		\bottomrule
	\end{tabular}

@@

Another category that may also be non-overt is the copula.  The primary strategy for copula constructions in Kazakh is the use of a defective verb \kazakh{е-}.  In the present tense, all but third person copulas have realised forms (e.g., \kazakh{Мен үйде\textbf{мін}.} \gloss{I'm at home.}), but the third person copula does not surface (e.g., \kazakh{Ол үйде.} \gloss{S/he's at home.}).

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{cop} \\
                                \gloss{Aygül} \& \gloss{student} \& \gloss{is} \\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx }\label{fig:compound}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \gloss{Aygül} \& \gloss{student} \& \gloss{was}\\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& Азаматтың \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{np.gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \gloss{Aygül} \& \gloss{Azamat} \& \gloss{where} \& \gloss{being} \& \gloss{knows not} \\
                        \end{deptext}
                        \depedge{5}{1}{\udlabel{subj}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop3}
        \end{subfigure}

@@

	\begin{dependency}[theme = simple, font = \small]
	   \begin{deptext}[column sep=0.08cm]
%               1         2         3          4              5                6              7                8           9             10
             Барлық \& адамдар \& тумысынан \& азат      \& және         \& қадір-қасиеті \& мен          \& құқықтары  \& тең       \& болып     \& дүниеге \& келеді \\
                    \&         \&           \& \gmk{adj} \& \gmk{cnjcoo} \& \gmk{n}       \& \gmk{cnjcoo} \& \gmk{n.pl} \& \gmk{adj} \& \gmk{cop} \& \& \\
                    \&         \&           \& \gloss{free} \& \gloss{and}  \& \gloss{dignity}  \&  \gloss{and} \& \gloss{rights} \&  \gloss{equal} \& \gloss{being} \& \& \\
		\end{deptext}
		\depedge{9}{4}{\udlabel{conj}}
		\depedge{9}{5}{\udlabel{cc}}
		\depedge{8}{6}{\udlabel{conj}}
		\depedge{8}{7}{\udlabel{cc}}
		\depedge{9}{8}{\udlabel{subj}}
		\depedge{10}{9}{\udlabel{cop}}
		\depedge{12}{10}{\udlabel{advcl}}
	\end{dependency}
	\caption{Coordination}\label{fig:coord}

@@

	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				көрші \& елдер \\
				\gmk{n.attr} \& \gmk{n.pl.nom} \\
				\gloss{neighbour} \& \gloss{countries} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{compound}}
		\end{dependency}
		\caption{Compounding: an attributive nominal depending on a nominal.}\label{fig:compound}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				әлем \& чемпионаты \\
				\gmk{n.nom} \& \gmk{n.px3sg.nom} \\
				\gloss{world} \& \gloss{championship} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{nmod}}
		\end{dependency}
		\caption{Indefinite genitive: An indefinite genitive depending on a nominal.}\label{fig:indefgen}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				Иранның \& экономиясы \\
				\gmk{n.gen} \& \gmk{n.px3sg.nom} \\
				\gloss{Iran's} \& \gloss{economy} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{nmod}}
		\end{dependency}
		\caption{Definite genitive: A definite genitive depending on a nominal.}\label{fig:defgen}
	\end{subfigure}

@@

\begin{table}[htbp]
	\centering
	\begin{tabular}{llrr}
		\toprule
			\textbf{Features}       & \textbf{Algorithm} &\textbf{LAS} [range] & \textbf{UAS} [range] \\
		\midrule
			surface                & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma          & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma+POS      & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
			surface+lemma+POS+MSD  & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
		\bottomrule
	\end{tabular}
	\caption{Preliminary parsing results from MaltParser using different models. The numbers in brackets denote the upper and lower bounds found during cross validation. Adding structural features to the model substantially improves the performance of the parser, although we find no improvement in using full morphosyntactic description (MSD) over using simply the first part-of-speech (POS) tag.}
	\label{table:eval}
\end{table}

>>>>

%%%%% Stuff for this paper %%%%%
\usepackage{tikz-dependency}
\newcommand{\gmk}[1]{{\rm {\sc \texttt{#1}}}}
\newcommand{\kazakh}[1]{{\em #1}}
\newcommand{\gloss}[1]{`#1'}
\newcommand{\tgloss}[1]{{\em #1}}
\newcommand{\tag}[1]{\texttt{‹#1›}}
%\newcommand{\udtag}[1]{\texttt{@#1}}
\newcommand{\udtag}[1]{{\ll \textsc{#1}}}
\newcommand{\udlabel}[1]{{\udfont #1}}
\newcommand{\tilda}{{\fsfont ∼}}
%\newcommand{\corrauthor}[1]{\footnotemark\footnotetext{hargle}}

@@

\begin{table}[htbp]
	\centering
	\caption{The default (first line of each) and ``zero-derived'' uses of nominals, adjectives, and adverbs}\label{tab:zeroderiv}
	\begin{tabular}{lp{8em}p{22em}}
		\toprule
			\textbf{Category} & \textbf{Function} & \textbf{Example} \\
		\midrule
			Nominals & Substantive\newline
							Attributive\newline
							Adverbial
						& hargle\newline 
							jargal\newline
							бір рет \gloss{once} \\
			Adjectives & Attributive\newline
							Substantive\newline
							Adverbial
						& \kazakh{\textbf{жақсы} үй} \gloss{nice house} \newline
						\kazakh{ең \textbf{жақсы}лары} \gloss{\textbf{the nice}st \textbf{ones}} \newline
						\kazakh{оны \textbf{жақсы} танимын} \gloss{I know him/her \textbf{well}.}
						\\
			Adverbs & Adverbial
						& 
						\\
		\bottomrule
	\end{tabular}

@@

Another category that may also be non-overt is the copula.  The primary strategy for copula constructions in Kazakh is the use of a defective verb \kazakh{е-}.  In the present tense, the verb itself does not surface, but agreement morphology surfaces (cliticised to the previous word) in all but the third person forms (e.g., \kazakh{Мен үйде\textbf{мін}.} \gloss{I'm at home.} versus \kazakh{Ол үйде.} \gloss{S/he's at home.}).  The defective copula verb also surfaces in the recent past tense (e.g., \kazakh{Мен үйде еді\textbf{м}.} \gloss{I was at home.}, \kazakh{Інім үйде еді.} \gloss{My younger brother was at home.}).  Of particular interest are non-finite forms of the copula.  Copula clauses can be attributive, as in \kazakh{үйі үлкен жігіт}

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{cop} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx }\label{fig:compound}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.3cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& Азаматтың \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{np.gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{Azamat} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \depedge{5}{1}{\udlabel{subj}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{xxx.}\label{fig:cop3}
        \end{subfigure}

@@

	\begin{dependency}[theme = simple, font = \small]
	   \begin{deptext}[column sep=0.08cm]
%               1         2         3          4              5                6              7                8           9             10
             Барлық \& адамдар \& тумысынан \& азат      \& және         \& қадір-қасиеті \& мен          \& құқықтары  \& тең       \& болып     \& дүниеге \& келеді \\
                    \&         \&           \& \gmk{adj} \& \gmk{cnjcoo} \& \gmk{n}       \& \gmk{cnjcoo} \& \gmk{n.pl} \& \gmk{adj} \& \gmk{cop} \& \& \\
                    \&         \&           \& \tgloss{free} \& \tgloss{and}  \& \tgloss{dignity}  \&  \tgloss{and} \& \tgloss{rights} \&  \tgloss{equal} \& \tgloss{being} \& \& \\
		\end{deptext}
		\depedge{9}{4}{\udlabel{conj}}
		\depedge{9}{5}{\udlabel{cc}}
		\depedge{8}{6}{\udlabel{conj}}
		\depedge{8}{7}{\udlabel{cc}}
		\depedge{9}{8}{\udlabel{subj}}
		\depedge{10}{9}{\udlabel{cop}}
		\depedge{12}{10}{\udlabel{advcl}}
	\end{dependency}
	\caption{Coordination}\label{fig:coord}

@@

	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				көрші \& елдер \\
				\gmk{n.attr} \& \gmk{n.pl.nom} \\
				\tgloss{neighbour} \& \tgloss{countries} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{compound}}
		\end{dependency}
		\caption{Compounding: an attributive nominal depending on a nominal.}\label{fig:compound}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				әлем \& чемпионаты \\
				\gmk{n.nom} \& \gmk{n.px3sg.nom} \\
				\tgloss{world} \& \tgloss{championship} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{nmod}}
		\end{dependency}
		\caption{Indefinite genitive: An indefinite genitive depending on a nominal.}\label{fig:indefgen}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.3\textwidth}
		\centering
		\begin{dependency}[theme = simple, font = \small]
		   \begin{deptext}[column sep=0.3cm]
				Иранның \& экономиясы \\
				\gmk{n.gen} \& \gmk{n.px3sg.nom} \\
				\tgloss{Iran's} \& \tgloss{economy} \\
			\end{deptext}
			\depedge{2}{1}{\udlabel{nmod}}
		\end{dependency}
		\caption{Definite genitive: A definite genitive depending on a nominal.}\label{fig:defgen}
	\end{subfigure}

@@

\begin{table}[htbp]
	\caption{Preliminary parsing results from MaltParser using different models. The numbers in brackets denote the upper and lower bounds found during cross validation. Adding structural features to the model substantially improves the performance of the parser, although we find no improvement in using full morphosyntactic description (MSD) over using simply the first part-of-speech (POS) tag.}
	\centering
	\begin{tabular}{llrr}
		\toprule
			\textbf{Features}       & \textbf{Algorithm} &\textbf{LAS} [range] & \textbf{UAS} [range] \\
		\midrule
			surface                & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma          & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma+POS      & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
			surface+lemma+POS+MSD  & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
		\bottomrule
	\end{tabular}
	\label{table:eval}
\end{table}
